---
title: "wage gap in Spain"
author: "sara navarro medina"
date: "25/01/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Average Salary data by occupation
Data https://ine.es/jaxiT3/files/t/es/csv_bdsc/10916.csv?nocab=
- Occupation group
- Sex/Gender gap
- Period (years)
- Total

## Average salaries data by autonomy
Data: https://www.ine.es/jaxiT3/files/t/es/csv_bdsc/13930.csv?nocab=1
- Type of day
- Decile
- Autonomous Communities and Cities
- Period (in years)
- Total

```{r message=FALSE, warning=FALSE}
if (!"gghighlight" %in% installed.packages()) {install.packages("gghighlight")} # to highlight lines
if (!"viridis" %in% installed.packages()) {install.packages('viridis')}     # Color palette
if (!"ggthemes" %in% installed.packages()) {install.packages('ggthemes')}    # Themes
if (!"mapSpain" %in% installed.packages()) {install.packages('mapSpain')}    # Spain maps
if (!"tmap" %in% installed.packages()) {install.packages('tmap') }       # Draw maps
if (!"cartography" %in% installed.packages()) {install.packages('cartography')} # Draw mapas
if (!"cartography" %in% installed.packages()) {install.packages('cartography')} # Draw mapas
```

```{r message=FALSE, warning=FALSE}
# Libraries
library(tidyverse)
library(lubridate)   # Date handling
library(scales)      # Format dates
library(gghighlight) # To highlight lines
library(viridis)     # Color palette
library(ggthemes)    # Themes
library(sf)          # Handle sf objects
library(mapSpain)    # Spain maps
library(tmap)        # Draw mapas
library(cartography) # Draw mapas
```

### Chart template
```{r}
theme_a <- function(base_size = 9,
                    base_family = "sans"
                           )
    {
 
    tema <-
        theme_bw(base_size=base_size) +
          
        theme(legend.position="right") +
        theme(legend.text = element_text(size=base_size+2,family = base_family)) +
        theme(plot.title=element_text(size=base_size+4, 
                                      vjust=1.25, 
                                      family=base_family, 
                                      hjust = 0.5
                                      )) +
        
        theme(plot.subtitle=element_text(size=base_size+2, family = base_family,  hjust = 0.5))  +
        theme(text = element_text(size=base_size+2,family = base_family)) +
        theme(axis.text.x=element_text(size=base_size+2,family = base_family)) +
        theme(axis.text.y=element_text(size=base_size+2, family = base_family)) +
          
        theme(axis.title.x=element_text(size=base_size+2, vjust=0, family = base_family)) +
        theme(axis.title.y=element_text(size=base_size+2, vjust=1.25, family = base_family)) +
        theme(plot.caption=element_text(size=base_size-1, family = base_family)) +
        theme(strip.text = element_text(size=base_size+1, family = base_family)) +
        theme(strip.text.x = element_text(size=base_size+1, family = base_family)) +
        theme(strip.text.y = element_text(size=base_size+1, family = base_family)) 
 
    return (tema)
}
```

### Correct wrong data
```{r}
locale(date_names = "es", date_format = "%AD", time_format = "%AT",
  decimal_mark = ",", grouping_mark = ".", tz = "UTC",
  encoding = "UTF-8", asciify = FALSE)

salarios_ocup <- read_csv2 ("https://ine.es/jaxiT3/files/t/es/csv_bdsc/10916.csv?nocab=")
```
There are 27 wrong rows. We look for them in the file and we see that there is an occupation group that separates the text with ; and the parser interprets it as a separator. we fix it like this
- We download the file
- We change the ; for ,
- We read the corrected file

```{r}
# download tmp.csv file next to this notebook
download.file("https://ine.es/jaxiT3/files/t/es/csv_bdsc/10916.csv?nocab=", "tmp.csv")
# reads the text
tx <- readLines("tmp.csv")
# replace text
tx <- gsub("; profesionales de apoyo",", profesionales de apoyo", tx)
# create file with text tx
writeLines(tx, "newtmp.csv")

salarios_ocup <- read_csv2 ("newtmp.csv",
                            na = "..",
                            col_names = TRUE,
                            cols(`Grupos de ocupación` = col_character(), 
                                `Sexo/Brecha de género` = col_character(),
                                 Periodo = col_date(format('%Y')),
                                 Total = col_number  ()))

names(salarios_ocup)
print (unique(sort(salarios_ocup$`Grupos de ocupación`)))
```

### Prepare data
- In some rows the value of Total is negative (it must be an error) make all the values ​​of that column positive using the abs() function
- The table in the Sex/Gender gap column is widened to facilitate calculations
- The difference in wages between men and women will be calculated in units of 1000 euros (hint: you will have to widen the table)
- The salary gap will be calculated 100 - column `Ratio of women to men` (percentage of women's salary compared to men's calculated as women's salary * 100/ Men's salary)
- The names of the occupation groups will be simplified because they are very long
  - Separate the code, the text. A little help, extract( `Occupation groups`, c("code","occupation"), "([[:alnum:]]+)([ [:alnum:]]*)")
  - The name will be simplified. A little help on how to make a dataframe in the code with other names. You can join the salaries_occupation data with this dataframe and you will have the simplified names accessible
  - Delete the occupation group "J" that has incomplete values
```{r}

salarios_diff <- salarios_ocup %>%
                mutate(Total = abs(Total)) %>% # Remove negative values
                pivot_wider (names_from = `Sexo/Brecha de género`, values_from = Total ) %>% # ensanchamos Sexo/Brecha de género
                mutate (diff = (Hombres - Mujeres)/1000) %>% #calculate difference between men and women
                mutate (brecha = 100 - `Cociente mujeres respecto a hombres`) %>% #calculate gap
                extract( `Grupos de ocupación`, c("code","ocup"), "([[:alnum:]]+)([ [:alnum:]]*)") %>% # separate occupation code
                filter (code != "J") # filter group J


code <- c("A","B","C","D","E","F","G","H","I","J","K","L","M","N","O","P","Total")
ocup_simple <- c( "Directores y gerentes",
                  "Científicos, salud y la enseñanza",
                  "Otros profesionales científicos",
                  "Profesionales de apoyo",
                  "Administrativos",
                  "Atención al público",
                  "Servicios de restauración y comercio",
                  "Servicios de salud y cuidados",
                  "Servicios de seguridad",
                  "Cualificados sector agrario",
                  "Cualificados sector construcción",
                  "Cualificados sector industrial",
                  "Operadores de instaladores",
                  "Conductores y maquinistas",
                  "No cualificados sector servicios",
                  "Peones",
                  "Total")
ocup_nombre <- tibble(code ,
                      ocup_simple)


# the occupied column is removed and a left join is made with ocup_nombre
salarios_diff <- salarios_diff %>%
                     select(-ocup) %>%
                     left_join(ocup_nombre, by = "code")


salarios_diff <- salarios_diff[, c(1, 8, 2, 3, 4, 5, 6, 7)] # change order columns

print(salarios_diff)
```

# Evolution of the wage gap between women and men by occupation group
```{r fig.width=9, fig.height=6}
# Heatmap
max_fecha = max(salarios_diff$Periodo)
df_order <- salarios_diff %>% 
            filter (Periodo == max_fecha) %>%
            arrange (brecha) 
orden_ocup <- unique(df_order$ocup_simple)

df <- filter(salarios_diff, code != "Total")

df$ocup_simple <- factor(df$ocup_simple,levels=orden_ocup)

# Heatmap
ggplot(data = df, aes(x = Periodo, y = ocup_simple, fill=brecha)) +
  geom_tile()+
  geom_text(aes(label = round(brecha,1)), color="white",
          size =3,  hjust=.5, vjust=.5 ) +
  scale_fill_gradient(low = "steelblue", high = "red4")+
  labs(x = "Año", fill = "Porcentaje de la\nBrecha\nSalarial",
       title = "Evolución del porcentaje de la brecha salarial de la mujer respecto del hombre\n por grupo de ocupación",
       caption = "Fuente: INE") + 
  scale_x_date(date_breaks = "1 year", 
               date_labels = "%y",
               expand =  c(0,0),
               sec.axis = dup_axis()) +
  theme_a() +
  theme (axis.title.y=element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())
```

```{r fig.width=8, fig.height=6}
# line chart
df$ocup_simple <- factor(df$ocup_simple, levels = ocup_nombre$ocup_simple)

ggplot() +
  geom_step(data = df, aes(x = Periodo, y = brecha, color = ocup_simple)) +
  geom_line(data = df, aes(x = Periodo, y = brecha), 
                                      color="grey", size = 1,
                                      alpha = 0.5) +
  labs(x = "Año", y = "Brecha Salarial de la mujer respecto del hombre",
       title = "Evolución del porcentaje de la brecha Salarial de la mujer respecto del hombre por grupo de ocupación",
       caption = "Fuente: INE") +
  guides(color = "none") +
  scale_color_viridis (option = 'plasma', discrete =TRUE,
                       direction =-1,
                       begin=0.2, end=0.8) + # Paleta contínua, uso discreto
  scale_y_continuous (labels=scales::percent_format(scale = 1,accuracy = 1L),
                      limits=c (0, 50),
                      breaks = seq(0, 40, by = 20)) + 
  scale_x_date (limits= c(min(df$Periodo, na.rm = TRUE)-365, max(df$Periodo, na.rm = TRUE)+(365)),
                breaks = c(min(df$Periodo), max(df$Periodo)),
                labels = date_format("%Y")) +  # Alargamos el eje x
  facet_wrap(~ ocup_simple, ncol = 3
             ) +
  theme_a()
```

# Evolution of the difference in average wages by occupation between men and women
```{r fig.width=9, fig.height=6}
# Heatmap
ggplot(data = df, aes(x = Periodo, y = ocup_simple, fill=diff)) +
  geom_tile()+
  geom_text(aes(label = round(diff,1)), color="white",
          size =3,  hjust=.5, vjust=.5 ) +
  scale_fill_gradient(low = "red4", high = "steelblue")+
  labs(x = "Año", fill = "Diferencia\nsalarial\nen miles",
       title = "Evolución de la diferencia de salarios medios, en miles, por ocupación entre hombres y mujeres",
       caption = "Fuente: INE") + 
  scale_x_date(date_breaks = "1 year", 
               date_labels = "%y",
               expand =  c(0,0),
               sec.axis = dup_axis()) +
  theme_a() +
  theme (axis.title.y=element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())
```

```{r fig.width=7, fig.height=6}
# Line chart
ggplot() +
  geom_step(data = df, aes(x = Periodo, y = diff, color = ocup_simple)) +
  geom_line(data = df, aes(x = Periodo, y = diff), 
                                      color="grey", size = 1,
                                      alpha = 0.5) +
  labs(x = "Año", y = "Diferencia salarial en miles",
       title = "Evolución de la diferencia de salarios medios, en miles, por ocupación entre hombres y mujeres",
       caption = "Fuente: INE") +
  guides(color = "none") +
  scale_color_viridis (option = 'plasma', discrete =TRUE,
                       direction =-1,
                       begin=0.2, end=0.8) + # Paleta contínua, uso discreto
  scale_y_continuous (limits=c (0, 30),
                      breaks = seq(0, 30, by = 10)) +
  scale_x_date (limits= c(min(df$Periodo, na.rm = TRUE)-365, max(df$Periodo, na.rm = TRUE)+(365)),
                breaks = c(min(df$Periodo), max(df$Periodo)),
                labels = date_format("%Y")) +  # Alargamos el eje x
  facet_wrap(~ ocup_simple, ncol = 3
             ) +
  theme_a()
```

# Variation of the salary gap between 2009 and 2019 by occupation group
```{r fig.width=7, fig.height=8}
# Dumbbell
df3 <- df[,c("ocup_simple","Periodo","brecha")]
df3 <- filter(df3, Periodo == "2019-01-01" | Periodo == "2009-01-01")
df3 <- df3 %>% pivot_wider(names_from = Periodo, values_from = brecha)
df3 <- df3 %>% rename("2019" = "2019-01-01", "2009" = "2009-01-01")

df4 <- df3 %>% mutate(diff = `2019`- `2009`) %>%
       mutate (ajuste_color = ifelse(diff >0, "Incremento", "Descenso"),
               ajuste_text = ifelse(diff >0, -0.2,1.2),
               ajuste_num = ifelse(diff > 0, 1.2,-0.2))
ggplot(data = df4) +
        
  geom_segment(aes(x = `2009`, 
                 xend = `2019`, 
                 y = reorder(ocup_simple,`2019`), 
                 yend = ocup_simple,
                 color = ajuste_color),
                 size=1,
                 arrow = arrow(length = unit(0.20,"cm"),
                               ends = "last",
                               type = "closed")) +
  geom_text(aes(x=`2009`, 
                y = reorder(ocup_simple,`2019`), 
                color = ajuste_color, 
                label = round(diff,1),
                hjust = 0),
                size = 4, vjust = -1, show.legend = FALSE) +
  geom_text(aes(x=`2019`, 
                y= reorder(ocup_simple,`2019`),
                color = ajuste_color,
                label = ocup_simple, 
                hjust = ajuste_text),
                size=4, vjust = 0, show.legend = FALSE) +
  labs(title = "Variación del porcentaje de la brecha salarial entre el 2009 y 2019\npor grupo de ocupación",
                caption = "Fuente: INE") +
  scale_colour_manual('', values = c('Descenso'='steelblue4', 'Incremento'='red4')) +
  scale_x_continuous (limit = c(0,35),
                      breaks = seq(0, 40, by = 10),
                      expand =  c(0,0),
                      sec.axis = dup_axis()) +
  scale_y_discrete ( expand =  c(0.09,0) ) +

  theme_a() +
  theme (legend.position = "top",
         axis.title.y = element_blank(),
         axis.text.y = element_blank(),
         axis.title.x = element_blank(),
         axis.ticks.y = element_blank(),
         panel.grid.major.y = element_blank(),
         panel.grid.major.x = element_line(colour = "gray"),
         plot.margin = margin(1, 1, 1,1, "cm"))
```

# Distribution of the wage gap by occupation group
```{r fig.width=7, fig.height=6}
# Histogram
ggplot(df) + 
  geom_histogram( aes(x = brecha, group = Periodo),
                 fill='cornflowerblue',
                 color = "white",
                 bins = 30)  +
  labs(title = "Evolución de la tasa de paro por edad",
       caption = "Fuente: INE") +
  scale_x_continuous (labels=scales::percent_format(scale = 1,accuracy = 1L),
                    #breaks =seq(0, 30, by = 10)) +
                   breaks =seq(0, max(df$brecha),
                               by = round(max(df$brecha)/min(df$brecha)+5,-1)))+
  guides(color = "none")+
  facet_wrap(~ ocup_simple, ncol=3)+
  theme_a() +
  theme (axis.title.y=element_blank(),
         axis.title.x=element_blank() ) 
```

# Relationship between wage gap VS salary difference between men and women by occupation group
```{r fig.width=9, fig.height=7}
# Dual axis line chart
ggplot( ) +
   geom_line( data= df, 
              aes(x = Periodo, y=brecha , color = "Brecha")) +
   geom_line( data = df,
              aes(x = Periodo, y=diff*2, color = "Diferencia")) + 
   labs(title = "Evolución entre el porcentaje de la brecha salarial\nvs.\ndiferencia salarial, en miles, entre hombres y mujeres por grupo de ocupación",
        x = "Años")+
   scale_y_continuous(name = "porcentaje de brecha salarial",
                      labels=scales::percent_format( scale = 1,accuracy = 1L), 
                         sec.axis = sec_axis(~./2, name="Diferencia salarial en miles")) + # Divide by 2
   scale_colour_manual('', values = c("Diferencia"='steelblue4', "Brecha"='red4')) +
   facet_wrap(~ ocup_simple, ncol=3) +
   theme_a() +
   theme( legend.position = "top",
          axis.title.y = element_text(color = "red4", size = 14),
          axis.title.y.right = element_text(color = "steelblue4", size = 14),
          axis.text.y = element_text(color = "red4"),
          axis.text.y.right = element_text(color = "steelblue4")
          )
```

# Map of average salaries by autonomies
Data: https://www.ine.es/jaxiT3/files/t/es/csv_bdsc/13930.csv?nocab=1
Space resources:
- Canarias: https://congosto.github.io/mapas/recintos_autonomicas_inspire_canarias_wgs84.json
- Peninsula and baleares: https://congosto.github.io/mapas/recintos_autonomicas_inspire_peninbal_etrs89.json
- Translation of INE names to ISO names: https://www.ine.es/jaxiT3/files/t/es/csv_bdsc/13930.csv?nocab=1
- Library mapspain

```{r}
# name translation resource
url_names_spain_region <- "https://raw.githubusercontent.com/montera34/escovid19data/master/data/original/shapes/spain_regions_utm.csv"
names_spain_region <- read_csv2(url_names_spain_region)
print (names_spain_region)

# reading the data file and settings
salarios_ccaa <- read_csv2 ("https://www.ine.es/jaxiT3/files/t/es/csv_bdsc/13930.csv?nocab=1",
                            na = "..")
 salarios_ccaa <- salarios_ccaa %>% 
                rename ( ccaa= `Comunidades y Ciudades Autonómas`) # chane variablee name
salarios_ccaa$ccaa <-  str_replace (salarios_ccaa$ccaa,"[0-9]+ ", "")  # Remove autonomy code

# We filter data to have the values of the salaries of the last year
salarios_ccaa <- salarios_ccaa %>% 
                         filter (`Tipo de jornada` == "Jornada a tiempo completo" &
                         Decil =="Total decil" &
                         ccaa != "Total Nacional" &
                         Periodo == max(Periodo)) 

# Union of data and normalized names
salarios_ccaa_nor <- left_join(names_spain_region,salarios_ccaa, by = c( "INE_name" = "ccaa"))
print (salarios_ccaa_nor)

# Read spatial files ffrom península and Canarias
peninbal <- "https://congosto.github.io/mapas/recintos_autonomicas_inspire_peninbal_etrs89.json"
canarias <- "https://congosto.github.io/mapas/recintos_autonomicas_inspire_canarias_wgs84.json"
map_p_sf <- st_read(peninbal)
map_c_sf <- st_read(canarias)
# reading the complete map of Spain
mapSpain_sf <- esp_get_ccaa()
```

# Map of average salaries
```{r fig.height=8, fig.width=10, message=FALSE, warning=FALSE}
map_salarios_ccaa_nor <- left_join (mapSpain_sf, salarios_ccaa_nor , by = c( "ine.ccaa.name" = "INE_name"))

ggplot(data = map_salarios_ccaa_nor) +
     geom_sf(aes(geometry = geometry,fill = Total), color ="#6C96A6") + 
     geom_sf_text (aes(label = round(Total,1)),
                   stat = "sf_coordinates",
                    size = 3) +
    labs(title = paste ("Salario medio en 2020"),
    fill ="Salario medio") +

    scale_fill_gradient (low = "#E1F9FB",high = "#1882AD",
                        guide = guide_legend()) +
    theme_map() +
    theme(plot.title=element_text(size='14',face="bold"),
          legend.title = element_text(size=11),
          legend.text = element_text(size=10),
          strip.text = element_text(size=9),
          legend.position = "right")
```